---
title: "DGEobject_DESq"
output: html_document
date: "2024-05-08"
---

# load libraries and set project directory 
```{r}
# load relevant libraries 
library(plyr)
library(reshape2)
library(dplyr)
library(stringr)
library(readxl)
library(readr)
library(magrittr)
library(ggplot2)
library(edgeR)
library(DESeq2)
library(limma)
library(Glimma)
library(RColorBrewer)
library(IRanges)
library(GenomicFeatures)

# set project directory
projectDir <- "/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/readCounts_analysis/rawData"

# set path for data tables  
import_brain_f <- file.path(projectDir, "rawCounts_brain_f.csv")
import_brain_m <- file.path(projectDir, "rawCounts_brain_m.csv")
import_heart_f <- file.path(projectDir, "rawCounts_heart_f.csv")
import_heart_m <- file.path(projectDir, "rawCounts_heart_m.csv")
import_liver_f <- file.path(projectDir, "rawCounts_liver_f.csv")
import_liver_m <- file.path(projectDir, "rawCounts_liver_m.csv")
import_lung_f <- file.path(projectDir, "rawCounts_lung_f.csv")
import_lung_m <- file.path(projectDir, "rawCounts_lung_m.csv")
import_muscle_f <- file.path(projectDir, "rawCounts_muscle_f.csv")
import_muscle_m <- file.path(projectDir, "rawCounts_muscle_m.csv")
import_placenta_f <- file.path(projectDir, "rawCounts_placenta_f.csv")
import_placenta_m <- file.path(projectDir, "rawCounts_placenta_m.csv")
import_metaData <- file.path(projectDir, "metaData.csv")
input_genome_gff3_file <- file.path(projectDir, "gencode.v29.chr_patch_hapl_scaff.annotation.gff3")
```

# import raw readCounts
```{r}
# import the counts tables
## brainF
brainF <- read_delim(file = import_brain_f,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

## brainM
brainM <- read_delim(file = import_brain_m,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

## heartF
heartF <- read_delim(file = import_heart_f,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

## heartM
heartM <- read_delim(file = import_heart_m,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

## liverF
liverF <- read_delim(file = import_liver_f,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

## liverM
liverM <- read_delim(file = import_liver_m,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

## lungF
lungF <- read_delim(file = import_lung_f,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

## lungM
lungM <- read_delim(file = import_lung_m,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

## muscleF
muscleF <- read_delim(file = import_muscle_f,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

## muscleaM
muscleM <- read_delim(file = import_muscle_m,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

## placentaF
placentaF <- read_delim(file = import_placenta_f,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

## placentaM
placentaM <- read_delim(file = import_placenta_m,
                     col_names = TRUE,
                     delim = ",") %>%
  dplyr::select(., ensembl = ...1, everything()) %>%
  as.data.frame()

# import metaData table 
metaData <- read_delim(file = import_metaData,
                       col_names = TRUE,
                       delim = ",") %>%
  as.data.frame()
```

# import gencode
```{r}
# import the gencodev32_gff_inFile file
all_gencode_v29 <- rtracklayer::import(input_genome_gff3_file)
# this file contains more information than we need here
# subset out only the columns we need
gene_data <- data.frame(ensembl = all_gencode_v29@elementMetadata$gene_id,
                        hgnc_symbol = all_gencode_v29@elementMetadata$gene_name,
                        seqnames = all_gencode_v29@seqnames,
                        gene_type = all_gencode_v29@elementMetadata$gene_type)
# we're left with multiple identical rows, keep one each
gencode_v29_gene_id_symbol_chr_biotype <- gene_data %>%
  dplyr::distinct(., .keep_all = TRUE)
# which leaves us with the following dimensions where there is a unique combination of ensembl ID, hgnc symbol and chromosome
dim(gencode_v29_gene_id_symbol_chr_biotype) # 64837 4
# this does however, leave us with duplicate ensembl IDs
dim(gencode_v29_gene_id_symbol_chr_biotype[!duplicated(gencode_v29_gene_id_symbol_chr_biotype$ensembl), ]) # 64792 4
# after generating a table to see the duplicates I can see that this is an X/Y chromosome issue. All genes are assigned to both chromosomes
duplicate_rows <- gencode_v29_gene_id_symbol_chr_biotype[duplicated(gencode_v29_gene_id_symbol_chr_biotype$ensembl) | duplicated(gencode_v29_gene_id_symbol_chr_biotype$ensembl, fromLast = TRUE),]
# remove duplicates (we don't use the chromosome information for anything)
gencode <- gencode_v29_gene_id_symbol_chr_biotype[!duplicated(gencode_v29_gene_id_symbol_chr_biotype$ensembl), ]
```

# merge countTables using dplyr outer join
```{r}
# merge countsTables to create total rawCounts df
total_rawCounts <- dplyr::full_join(brainF, brainM, by = "ensembl") %>%
  dplyr::full_join(.,heartF, by = "ensembl") %>%
  dplyr::full_join(.,heartM, by = "ensembl") %>%
  dplyr::full_join(.,liverF, by = "ensembl") %>%
  dplyr::full_join(.,liverM, by = "ensembl") %>%
  dplyr::full_join(.,lungF, by = "ensembl") %>%
  dplyr::full_join(.,lungM, by = "ensembl") %>%
  dplyr::full_join(.,muscleF, by = "ensembl") %>%
  dplyr::full_join(.,muscleM, by = "ensembl") %>%
  dplyr::full_join(.,placentaF, by = "ensembl") %>%
  dplyr::full_join(.,placentaM, by = "ensembl") %>%
  tibble::column_to_rownames("ensembl")

# replace any NA values with zero
total_rawCounts[is.na(total_rawCounts)] <- 0

# subset gencode to only include protein coding genes
gencode_protein <- dplyr::filter(gencode, gene_type == "protein_coding")

# subset total rawCounts to only include protein coding genes
rawCounts_protein <- subset(total_rawCounts, rownames(total_rawCounts) %in% gencode_protein$ensembl)
```

# remove scattered samples (post normalisation)
```{r}
# remove SRR3191758 and SRR3191757
rawCounts_protein <- rawCounts_protein[,-101]
rawCounts_protein <- rawCounts_protein[,-98]
```

# correcting for batch effects
```{r}
# correct for batch effects before DESq2 normalisation and DGEobject creation 
rawCounts_protein <- limma::removeBatchEffect(rawCounts_protein)
```

# DESq2 normalisation
```{r}
# create DESeq2 matrix
DESq_rawCounts <- DESeqDataSetFromMatrix(countData = rawCounts_protein,
                                  colData = metaData,
                                  design = ~ tissue)

# estimate size factors for normalisation
dds <- estimateSizeFactors(DESq_rawCounts)

# get normalised counts
rawCounts_DESq <- counts(dds, normalized = TRUE)
```

# normalised vs unnormalised box plots
```{r}
# pre-plot process
dge_unnorm <- total_rawCounts
dge_norm <- rawCounts_DESq

# box plot of un-normalised expression distribution 
par(mfrow=c(1,2))
lcpm_unnorm <- cpm(dge_unnorm, log=TRUE)
boxplot(lcpm_unnorm, ylab="Log-cpm", las=2, col=blues9, main="Normalised data")

# box plot of normalised expression distribution 
par(mfrow=c(1,2))
lcpm_norm <- cpm(dge_norm, log=TRUE)
boxplot(lcpm_norm, ylab="Log-cpm", las=2, col=blues9, main="Normalised data")
```

# create DGEobject out of DESq2 normalised counts
```{r}
dgelist_DESq <- edgeR::DGEList(
  counts = rawCounts_DESq,
  samples = metaData,
  group = metaData$tissue,
  genes = rownames(rawCounts_DESq))
```

# cpm avergage by sex_tissue 
```{r}
# use the edgeR built in function `cpm()` to make a new table of cpm counts
cpms_sex <- edgeR::cpm(dgelist_DESq, log = FALSE)
# use the cpms dataframe to grab all the ensembl IDs for the genes
genes_to_average <- rownames(cpms_sex)
# create a new object for the mean cpms
# first step is to transform the matrix (flip it 90' so genes are now in the columns and samples are in the rows)
mean_sex_tissue_cpms <- t(cpms_sex) %>%
  # make sure you have a data frame
  data.frame() %>%
  # move the samplenames into a column (we need this to join the metadata table)
  tibble::rownames_to_column("sampleID") %>%
  # join the counts and the metadata table (but we only want the samplename and tissue columns for your data so we're subsetting using the square brackets)
  dplyr::left_join(., cpm_dgelist_DESq$samples[, c("sampleID", "tissue", "sex")], by = join_by(sampleID)) %>%
  # drop the samplename column again (it was added by the join)
  dplyr::select(., -sampleID) %>%
   # add a column
  dplyr::mutate(sex_tissue = paste(sex, tissue, sep = "_")) %>%
  # group by the column of interest. Here I've use the "sex_outcome" column but in your data it will be the "tissue" column
  group_by(sex_tissue) %>%
  # calculate the mean cpms of all genes (using the ensembl gene vector we made earlier)
  dplyr::summarise_at(vars(genes_to_average), mean)
```

# subset sex_tissue into female and male dataframes
```{r}
# subset female df
mean_female_cpms <- subset(mean_sex_tissue_cpms, startsWith(sex_tissue, "female_")) 
## export female DESq normalised cpm table 
write.csv(mean_female_cpms, file = file.path(projectDir, "mean_female_cpms_DESq.csv"))

# subset male df
mean_male_cpms <- subset(mean_sex_tissue_cpms, startsWith(sex_tissue, "male_"))
## export male DESq normalised cpm table 
write.csv(mean_male_cpms, file = file.path(projectDir, "mean_male_cpms_DESq.csv"))
```

