---
title: "male_tissueRank_DESq"
output: html_document
date: "2024-05-09"
---

# load libraries and set project directory 
```{r}
# load relevant libraries 
library(plyr)
library(reshape2)
library(dplyr)
library(stringr)
library(readxl)
library(readr)
library(magrittr)
library(ggplot2)
library(edgeR)
library(limma)
library(Glimma)
library(RColorBrewer)

# set project directory
projectDir <- "/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/readCounts_analysis/rawData"

# set path for male DESq normalised ncpm counts
import_mean_male_cpms_DESq <- file.path(projectDir, "mean_male_cpms_DESq.csv")
```

# import data tables 
```{r}
# import mean male cpms table
ncpms_male_DESq <- read_delim(file = import_mean_male_cpms_DESq,
                                   col_names = TRUE,
                                   delim = ",") %>%
  as.data.frame() %>%
  # remove the first column "...1" that;s input when the file is imported
  dplyr::select(., -...1)

# copy mean_ncpms for subsequent analysis 
ncpms_male_DESq_cp <- as.data.frame(ncpms_male_DESq)
```

# ranking normalised cpms by male_tissue
```{r}
# make calc_rank a function
calculate_rank <- function(x) {
  return(rank(-x))
}

# rank normalised cpms by male_tissue
ncpms_male_DESq[, paste0(names(ncpms_male_DESq)[-1], "_Rank")] <- apply(ncpms_male_DESq [, -1, drop = FALSE], 2, calculate_rank)

# create a df consistingof just '_Rank'
## convert sex_tissue from a column to row names
ncpms_male_DESq <- tibble::column_to_rownames(ncpms_male_DESq, var = "sex_tissue")
# create a df selecting just for _Rank
ranked_male_cpms_DESq <- data.frame(ncpms_male_DESq) %>%
  dplyr::select(matches("_Rank"))
```

# isolate all instances where placenta is ranked 6th
```{r}
# create a rank table
## create a df with 'geneid' in rows and samples in columns
ranked_male_DESq <- t(ranked_male_cpms_DESq) %>%
  data.frame() %>%
  ## convert 'geneid' from being row names into a column
  tibble::rownames_to_column("geneid")  
  
## remove "_Rank" from the end of the geneids
ranked_male_DESq <- tidyr::separate(
  ranked_male_DESq,
  col = geneid,
  into = c("geneid", "right"),
  sep = "_Rank") %>%
  mutate(geneid = str_remove(geneid, "_Rank")) %>%
  dplyr::select(-right)
  
## rename from sex_tissue to sex_tissue_rank
colnames(ranked_male_DESq) <- gsub("male_brain", "male_brain_rank", colnames(ranked_male_DESq))
colnames(ranked_male_DESq) <- gsub("male_heart", "male_heart_rank", colnames(ranked_male_DESq))
colnames(ranked_male_DESq) <- gsub("male_liver", "male_liver_rank", colnames(ranked_male_DESq))
colnames(ranked_male_DESq) <- gsub("male_lung", "male_lung_rank", colnames(ranked_male_DESq))
colnames(ranked_male_DESq) <- gsub("male_muscle", "male_muscle_rank", colnames(ranked_male_DESq))
colnames(ranked_male_DESq) <- gsub("male_placenta.", "male_placenta_rank", colnames(ranked_male_DESq))
  
# create ncpms table in the same format as rank table so they can be joined
## transpose the data frame to flip rows and columns (rows become columns and columns become rows)
nCPMs_male_DESq <- data.frame(ncpms_male_DESq_cp) %>%
  tibble::column_to_rownames(., var = "sex_tissue") %>%
  t() %>%
  data.frame() %>%
  ## convert 'geneids' from row names to into a column
  tibble::rownames_to_column("geneid")

## rename from sex_tissue to sex_tissue_ncpm
colnames(nCPMs_male_DESq) <- gsub("male_brain", "male_brain_ncpm", colnames(nCPMs_male_DESq))
colnames(nCPMs_male_DESq) <- gsub("male_heart", "male_heart_ncpm", colnames(nCPMs_male_DESq))
colnames(nCPMs_male_DESq) <- gsub("male_liver", "male_liver_ncpm", colnames(nCPMs_male_DESq))
colnames(nCPMs_male_DESq) <- gsub("male_lung", "male_lung_ncpm", colnames(nCPMs_male_DESq))
colnames(nCPMs_male_DESq) <- gsub("male_muscle", "male_muscle_ncpm", colnames(nCPMs_male_DESq))
colnames(nCPMs_male_DESq) <- gsub("male_placenta.", "male_placenta_ncpm", colnames(nCPMs_male_DESq))

# combine the rank and ncpm tables via dplyr::left_join
nCPMs_rank_male <- dplyr::left_join(ranked_male_DESq, nCPMs_male_DESq, copy = TRUE, by = join_by(geneid))
  
# select for when placenta_rank = 6 (placenta ncpm is ranked last for the given genes)
placenta_male_6_DESq <- subset(nCPMs_rank_male, male_placenta_rank == "6")
```

# filter for when nCPM > 1 for tissue rank 5
```{r}
# filter for where rank = 5 and cpm > 1
nCPM_rank_5_DESq <- dplyr::filter(placenta_male_6_DESq, male_brain_rank == "5" & male_brain_ncpm > 1 | male_heart_rank == "5" & male_heart_ncpm > 1 | male_liver_rank == "5" & male_liver_ncpm > 1 | male_lung_rank == "5" & male_lung_ncpm > 1 | male_muscle_rank == "5" & male_muscle_ncpm > 1)
```

# separate nCPM_rank_5_DESq into rank and ncpm tables respectively 
```{r}
# separate rank and ncpm into their own tables again to avoid introduction of geneid duplicates when the table is converted to long form. Will need to convert the rank and ncpm tables into long form individually and then join them together for nCPM(rank=5)/nCPM(rank=6) > 3 calculations. 

## separate rank into its own table
rank_5_DESq <- tibble::column_to_rownames(nCPM_rank_5_DESq, var = "geneid") %>%
 data.frame() %>%
  dplyr::select(matches("_rank"))

## separate ncpm into its own table
nCPM_5_DESq <- tibble::column_to_rownames(nCPM_rank_5_DESq, var = "geneid") %>%
 data.frame() %>%
  dplyr::select(matches("_ncpm"))
```

# convert rank and ncpm tables into one long form tables and stitch together for nCPM(rank=5)/nCPM(rank=6) > 3 calculation
```{r}
# convert rank_5_DESq into long form
## convert 'gemeids' from row names into a column
rank_5_DESq_long <- tibble::rownames_to_column(rank_5_DESq, var = "geneid")
## remove "_rank" from column names
colnames(rank_5_DESq_long) <- gsub("_rank", "", colnames(rank_5_DESq_long))
## convert into long form
rank_5_DESq_long <- tidyr::gather(rank_5_DESq_long, tissue, rank, male_brain, male_heart, male_liver, male_lung, male_muscle, male_placenta)
## craete a df where tissue rank = 5 to calculate the nCPM(rank=5)/nCPM(rank=6) ratio
rank_5_DESq_long_2 <- dplyr::filter(rank_5_DESq_long, rank == "5")

# convert nCPM_5_DESq into long form
## convert 'gemeids' from row names into a column
nCPM_5_DESq_long <- tibble::rownames_to_column(nCPM_5_DESq, var = "geneid")
## rename, remove "_cpm"
colnames(nCPM_5_DESq_long) <- gsub("_ncpm", "", colnames(nCPM_5_DESq_long))
  #convert into long form
nCPM_5_DESq_long <- tidyr::gather(nCPM_5_DESq_long, tissue, ncpm, male_brain, male_heart, male_liver, male_lung, male_muscle, male_placenta)

# join the rank and ncpm long form tables into one containing ncpms where tissue = rank 5 
nCPM_rank_5_DESq_long <- dplyr::left_join(rank_5_DESq_long_2, nCPM_5_DESq_long, by = c("geneid", "tissue"))
## rename ncpm to nCPM_5
colnames(nCPM_rank_5_DESq_long) <- gsub("ncpm", "nCPM_5", colnames(nCPM_rank_5_DESq_long))
```

# create a long df of just placenta_rank = 6 with geneids and ncpms for nCPM(rank=5)/nCPM(rank=6) calculation
```{r}
# create a long df with just the instances where placenta = 6
rank_placenta_long <- dplyr::filter(rank_5_DESq_long, rank == "6")

# join nCPM_5_DESq_long to rank_placenta_long to add ncpms for geneids where placenta rank = 6
placenta_6_nCPM <- dplyr::left_join(rank_placenta_long, nCPM_5_DESq_long, by = c("geneid", "tissue"))

# clean table 
## rename ncpms so column names are unique
colnames(placenta_6_nCPM) <- gsub("ncpm", "nCPM_6", colnames(placenta_6_nCPM))
## remove rank and tissue columns
placenta_6_nCPM <- dplyr::select(placenta_6_nCPM, -tissue, -rank)
```

#  filter for wherenCPM(rank=5)/nCPM(rank=6) > 3
```{r}
# join placenta_6_nCPM and nCPM_rank_5_DESq_long creating a df with both the nCPMs of rank 5 and 6
nCPM5_nCPM6_DESq_male <- dplyr::left_join(nCPM_rank_5_DESq_long, placenta_6_nCPM, by = "geneid")

# create a new column for nCPM5/nCPM6
nCPM5_nCPM6_DESq_male$nCPM_ratio <- nCPM5_nCPM6_DESq_male$nCPM_5 / nCPM5_nCPM6_DESq_male$nCPM_6

# select for nCPM_ratio > 3
placenta_m_depleted_DESq <- nCPM5_nCPM6_DESq_male[nCPM5_nCPM6_DESq_male$nCPM_ratio >3,]
```

# tidy and export placenta_m_depleted_DESq
```{r}
# replace column name 'geneid' with 'ensembl'
colnames(placenta_m_depleted_DESq) <- gsub("geneid", "ensembl", colnames(placenta_m_depleted_DESq))

# export placenta_m_depleted_DESq into project rawData for further analysis  
write.csv(placenta_m_depleted_DESq, file = file.path(projectDir, "DESq_placentaM_depleted.csv"))
```
