---
title: "TMM_heartF_rankAnalysis"
output: html_document
---

# load libraries and set project directory 
```{r}
# load relevant libraries 
library(plyr)
library(reshape2)
library(dplyr)
library(stringr)
library(readxl)
library(readr)
library(magrittr)
library(ggplot2)
library(edgeR)
library(limma)
library(Glimma)
library(RColorBrewer)

# set project directory
projectDir <- "/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/tissue_depletion/rawData"
```

# load data files
```{r}
## load mean female TMM normalised CPM data from termPlacenta_comparison analysis
ncpms_female_TMM <- read_delim(file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/readCounts_analysis/rawData/mean_female_cpms_TMM.csv"),
                                    col_names = TRUE,
                                    delim = ",") %>% 
  as.data.frame() %>%
  # remove the first column "...1" that;s input when the file is imported
  dplyr::select(., -...1)

# copy mean_cpms for subsequent analysis 
ncpms_female_TMM_cp <- as.data.frame(ncpms_female_TMM)
```

# ranking TMM normalised ncpms by female_tissue
```{r}
# make calc_rank a function
calculate_rank <- function(x) {
  return(rank(-x))
}

# rank normalised cpms by female_tissue
ncpms_female_TMM[, paste0(names(ncpms_female_TMM)[-1], "_Rank")] <- apply(ncpms_female_TMM [, -1, drop = FALSE], 2, calculate_rank)

# create a df consisting of just '_Rank'
## convert sex_tissue from a column to row names
ncpms_female_TMM <- tibble::column_to_rownames(ncpms_female_TMM, var = "sex_tissue")
# create a df selecting just for _Rank
ranked_female_cpms_TMM <- data.frame(ncpms_female_TMM) %>%
  dplyr::select(matches("_Rank"))
```

# isolate all instances where placenta is ranked 6th
```{r}
# create a rank table
## create a df with 'geneid' in rows and samples in columns
ranked_female_TMM <- t(ranked_female_cpms_TMM) %>%
  ## ensure it's in data frame structure
  data.frame() %>%
  ## convert 'geneid' from row names to a column
  tibble::rownames_to_column("geneid")  
  
## remove "_Rank" from the end of the geneids
ranked_female_TMM <- tidyr::separate(
  ranked_female_TMM,
  col = geneid,
  into = c("geneid", "right"),
  sep = "_Rank") %>%
  mutate(geneid = str_remove(geneid, "_Rank")) %>%
  dplyr::select(-right)
  
## rename from sex_tissue to sex_tissue_rank
colnames(ranked_female_TMM) <- gsub("female_brain", "female_brain_rank", colnames(ranked_female_TMM))
colnames(ranked_female_TMM) <- gsub("female_heart", "female_heart_rank", colnames(ranked_female_TMM))
colnames(ranked_female_TMM) <- gsub("female_liver", "female_liver_rank", colnames(ranked_female_TMM))
colnames(ranked_female_TMM) <- gsub("female_lung", "female_lung_rank", colnames(ranked_female_TMM))
colnames(ranked_female_TMM) <- gsub("female_muscle", "female_muscle_rank", colnames(ranked_female_TMM))
colnames(ranked_female_TMM) <- gsub("female_placenta.", "female_placenta_rank", colnames(ranked_female_TMM))
  
# create ncpms table in the same format as rank table so they can be joined
## transpose the data frame to flip the rows and columns (rows become columns and columns become rows)
nCPMs_female_TMM <- data.frame(ncpms_female_TMM_cp) %>%
  ## makes the column sex_tissue row names 
  tibble::column_to_rownames(., var = "sex_tissue") %>%
  t() %>%
  data.frame() %>%
  ## change geneids from row names to a column 
  tibble::rownames_to_column("geneid")

## rename to sex_tissue_cpm
colnames(nCPMs_female_TMM) <- gsub("female_brain", "female_brain_ncpm", colnames(nCPMs_female_TMM))
colnames(nCPMs_female_TMM) <- gsub("female_heart", "female_heart_ncpm", colnames(nCPMs_female_TMM))
colnames(nCPMs_female_TMM) <- gsub("female_liver", "female_liver_ncpm", colnames(nCPMs_female_TMM))
colnames(nCPMs_female_TMM) <- gsub("female_lung", "female_lung_ncpm", colnames(nCPMs_female_TMM))
colnames(nCPMs_female_TMM) <- gsub("female_muscle", "female_muscle_ncpm", colnames(nCPMs_female_TMM))
colnames(nCPMs_female_TMM) <- gsub("female_placenta.", "female_placenta_ncpm", colnames(nCPMs_female_TMM))

# combine the rank and ncpm tables via dplyr::left_join
nCPMs_rank_female_TMM <- dplyr::left_join(ranked_female_TMM, nCPMs_female_TMM, copy = TRUE, by = join_by(geneid))
  
# select for when brain_rank = 6 (brain ncpm is ranked last for the given genes)
heart_female_6_TMM <- subset(nCPMs_rank_female_TMM, female_heart_rank == "6")
```

# filter for when nCPM > 1 for tissue rank 5
```{r}
# filter for where rank = 5 and cpm > 1
nCPM_rank_5_TMM <- dplyr::filter(heart_female_6_TMM, female_brain_rank == "5" & female_brain_ncpm > 1 | female_lung_rank == "5" & female_lung_ncpm > 1 | female_liver_rank == "5" & female_liver_ncpm > 1 | female_muscle_rank == "5" & female_muscle_ncpm > 1 | female_placenta_rank == "5" & female_placenta_ncpm > 1)
```

# separate nCPM_rank_5_DESq into rank and ncpm tables respectively 
```{r}
# separate rank and ncpm into their own tables again to avoid introduction of geneid duplicates when the table is converted to long form. Will need to convert the rank and ncpm tables into long form individually and then join them together for nCPM(rank=5)/nCPM(rank=6) > 3 calculations. 
## separate rank into its own table
rank_5_TMM <- tibble::column_to_rownames(nCPM_rank_5_TMM, var = "geneid") %>%
 data.frame() %>%
  dplyr::select(matches("_rank"))

## separate ncpm into its own table
nCPM_5_TMM <- tibble::column_to_rownames(nCPM_rank_5_TMM, var = "geneid") %>%
 data.frame() %>%
  dplyr::select(matches("_ncpm"))
```

# convert rank and ncpm tables into one long form tables and stitch together for nCPM(rank=5)/nCPM(rank=6) > 3 calculation
```{r}
# convert rank_5_TMM into long form
## convert geneid from row names into a column
rank_5_TMM_long <- tibble::rownames_to_column(rank_5_TMM, var = "geneid")
## remove "_rank" from column names 
colnames(rank_5_TMM_long) <- gsub("_rank", "", colnames(rank_5_TMM_long))
## convert into long form
rank_5_TMM_long <- tidyr::gather(rank_5_TMM_long, tissue, rank, female_brain, female_heart, female_liver, female_lung, female_muscle, female_placenta)
## create a df where tissue rank = 5 to calculate the nCPM(rank=5)/nCPM(rank=6) ratio
rank_5_TMM_long_2 <- dplyr::filter(rank_5_TMM_long, rank == "5")

# convert cpm_5_TMM into long form
## convert geenids from row names into a column
nCPM_5_TMM_long <- tibble::rownames_to_column(nCPM_5_TMM, var = "geneid")
## remove "_ncpm" from column names 
colnames(nCPM_5_TMM_long) <- gsub("_ncpm", "", colnames(nCPM_5_TMM_long))
## convert into long form
nCPM_5_TMM_long <- tidyr::gather(nCPM_5_TMM_long, tissue, ncpm, female_brain, female_heart, female_liver, female_lung, female_muscle, female_placenta)

# join the rank and ncpm long form tables into one containing ncpms where tissue = rank 5 
nCPM_rank_5_TMM_long <- dplyr::left_join(rank_5_TMM_long_2, nCPM_5_TMM_long, by = c("geneid", "tissue"))
## rename ncpm to nCPM_5
colnames(nCPM_rank_5_TMM_long) <- gsub("ncpm", "nCPM_5", colnames(nCPM_rank_5_TMM_long))
```

# create a long df of just placenta_rank = 6 with geneids and ncpms for nCPM(rank=5)/nCPM(rank=6) calculation
```{r}
# create a long df with just the instances where placenta = 6
rank_heart_long <- filter(rank_5_TMM_long, rank == "6")

# join nCPM_5_TMM_long to rank_lung_long to add ncpms for geneids where placenta rank = 6
heart_6_nCPM <- dplyr::left_join(rank_liver_long, nCPM_5_TMM_long, by = c("geneid", "tissue"))

# clean table   
## rename ncpms so column anmes are unique
colnames(heart_6_nCPM) <- gsub("ncpm", "nCPM_6", colnames(heart_6_nCPM))
## remove rank and tissue columns as we only require ncpm(rank=6) for ratio calculation
heart_6_nCPM <- dplyr::select(heart_6_nCPM, -tissue, -rank)
```

# filter for where nCPM(rank=5)/nCPM(rank=6) > 3
```{r}
# join heart_6_nCPM and nCPM_rank_5_TMM_long creating a df with both the nCPMs of rank 5 and 6
nCPM5_nCPM6_TMM_female <- dplyr::left_join(nCPM_rank_5_TMM_long, heart_6_nCPM, by = "geneid")

# create a new column for nCPM5/nCPM6
nCPM5_nCPM6_TMM_female$nCPM_ratio <- nCPM5_nCPM6_TMM_female$nCPM_5 / nCPM5_nCPM6_TMM_female$nCPM_6

# select for nCPM_ratio > 3
heart_f_depleted_TMM <- nCPM5_nCPM6_TMM_female[nCPM5_nCPM6_TMM_female$nCPM_ratio >3,]
```

# tidy and export lung_f_depleted_TMM
```{r}
# replace column name 'geneid' with 'ensembl'
colnames(heart_f_depleted_TMM) <- gsub("geneid", "ensembl", colnames(heart_f_depleted_TMM))

# export heart_f_depleted_TMM into project rawData for further analysis 
write.csv(heart_f_depleted_TMM, file = file.path(projectDir, "TMM_heartF_depleted.csv"))
```
