---
title: "termM_sexSpecific_rankAnalysis"
output: html_document
date: "2024-07-29"
---

# load libraries and set project dir
```{r}
# load relevant libraries 
library(plyr)
library(reshape2)
library(dplyr)
library(stringr)
library(readxl)
library(readr)
library(magrittr)
library(ggplot2)
library(edgeR)
library(limma)
library(Glimma)
library(RColorBrewer)

# set project directory
projectDir <- "/Users/taylaalbertini/Desktop/R studio/honours_project/term_sex_specific_analysis/rawData"
```

# load data files
```{r}
## load mean female TMM normalised CPM data from termPlacenta_comparison analysis
mean_termF_ncpms_TMM <- read_delim(file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/termPlacenta_comparison/rawData/mean_female_cpms_termC_TMM.csv"),
                                    col_names = TRUE,
                                    delim = ",") %>% 
  as.data.frame() %>%
  # remove the first column "...1" that;s input when the file is imported
  dplyr::select(., -...1) %>% 
  ## filter data table to only include term counts - remove mid-gestation placenta in row 1
  dplyr::slice(-1) %>%
  tibble::column_to_rownames(., var = "sex_tissue") 
## make row name unique 
rownames(mean_termF_ncpms_TMM) <- "termF_placenta"


## load mean male TMM normalised CPM data from termPlacenta_comparison analysis 
mean_termM_ncpms_TMM <- read_delim(file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/termPlacenta_comparison/rawData/mean_male_cpms_termC_TMM.csv"),
                                    col_names = TRUE,
                                    delim = ",") %>% 
  as.data.frame() %>%
  # remove the first column "...1" that;s input when the file is imported
  dplyr::select(., -...1) %>% 
  ## filter data table to only include term counts - remove mid-gestation placenta in row 1
  dplyr::slice(-1) %>% 
  tibble::column_to_rownames(., var = "sex_tissue")
## make row name unique 
rownames(mean_termM_ncpms_TMM) <- "termM_placenta"

## combine the two mean counts tables into one
mean_term_ncpms_TMM <- rbind(mean_termF_ncpms_TMM, mean_termM_ncpms_TMM)
## check to see if any NA values were input 
sum(is.na(mean_term_ncpms_TMM))
## make a copy of mean_term_ncpms_TMM for downstream analysis
mean_term_ncpms_TMM_cp <- as.data.frame(mean_term_ncpms_TMM)
```

# ranking TMM normalised cpms by sex_tissue
```{r}
# make calc_rank a function
calculate_rank <- function(x) {
  return(rank(-x))
}

# rank normalised cpms by sex_tissue
mean_term_ncpms_TMM[, paste0(names(mean_term_ncpms_TMM)[-1], "_Rank")] <- apply(mean_term_ncpms_TMM [, -1, drop = FALSE], 2, calculate_rank)

## create the df filtering just for _Rank
ranked_term_ncpms_TMM <- data.frame(mean_term_ncpms_TMM) %>%
  dplyr::select(matches("_Rank"))
```

# isolate all instances where female term placenta is ranked 2nd
```{r}
# creating the rank table
  ## create a dataframe with geneid in rows and samples in columns
ranked_term_TMM <- t(ranked_term_ncpms_TMM) %>%
  data.frame() %>%
   ## change gene ensembls from being row names to its own column, matching the format of rankComparison_female_TMM
  tibble::rownames_to_column("geneid")  
  
  ### remove "_Rank" from the end of the geneids
  ranked_term_TMM <- tidyr::separate(
  ranked_term_TMM,
  col = geneid,
  into = c("geneid", "right"),
  sep = "_Rank") %>%
  mutate(geneid = str_remove(geneid, "_Rank")) %>%
  dplyr::select(-right)
  
# rename to sex_tissue_rank
colnames(ranked_term_TMM) <- gsub("termF_placenta", "termF_placenta_rank", colnames(ranked_term_TMM))
colnames(ranked_term_TMM) <- gsub("termM_placenta", "termM_placenta_rank", colnames(ranked_term_TMM))
    
# create the nCPMs table in the same format as rank table so they can be joined
## transpose the data frame, flips rows and columns (rows become columns and columns become rows)
nCPMs_term_TMM <- t(mean_term_ncpms_TMM) %>%
  data.frame() %>%
  ## change gene ensembls from being row names to its own column, matching the format of rankComparison_female_TMM
  tibble::rownames_to_column("geneid")

# rename to sex_tissue_cpm
colnames(nCPMs_term_TMM) <- gsub("termF_placenta", "termF_placenta_cpm", colnames(nCPMs_term_TMM))
colnames(nCPMs_term_TMM) <- gsub("termM_placenta", "termM_placenta_cpm", colnames(nCPMs_term_TMM))

# combine via left_join nCPMs_term_TMM and ranked_term_TMM
nCPMs_rank_male <- dplyr::left_join(ranked_term_TMM, nCPMs_term_TMM, copy = TRUE, by = join_by(geneid))
  
# select for when placenta_rank = 2
term_rank2_TMM <- subset(nCPMs_rank_male, termM_placenta_rank == "2")
```

# filter for when nCPM > 1 for tissue rank 1
```{r}
# filter for where rank = 5 and cpm > 1
nCPM_rank_1_TMM <- dplyr::filter(term_rank2_TMM, termF_placenta_rank == "1" & termF_placenta_cpm > 1)
```

# separate ranked_term_TMM and nCPMs_term_TMM into rank and cpm tables respectively 
```{r}
# separate rank and ncpm into their own tables again to avoid introduction of geneid duplicates when the table is converted to long form. Will need to convert the rank and ncpm tables into long form individually and then join them together for nCPM(rank=1)/nCPM(rank=2) > 3 calculations. 
## separate rank into its own table
rank_1_TMM <- tibble::column_to_rownames(nCPM_rank_1_TMM, var = "geneid") %>%
  data.frame() %>%
  dplyr::select(matches("_rank"))

## separate ncpm into its own table
nCPM_1_TMM <- tibble::column_to_rownames(nCPM_rank_1_TMM, var = "geneid") %>%
  data.frame() %>%
  dplyr::select(matches("_cpm"))
```

# convert rank and ncpm tables into one long form tables and stitch together for nCPM(rank=1)/nCPM(rank=2) > 3 calculation
```{r}
# convert cpm_1_TMM into long form
## convert rows into a column
nCPM_1_TMM_long <- tibble::rownames_to_column(nCPM_1_TMM, var = "geneid")
## rename, remove "_cpm"
colnames(nCPM_1_TMM_long) <- gsub("_cpm", "", colnames(nCPM_1_TMM_long))
#convert into long form
nCPM_1_TMM_long <- tidyr::gather(nCPM_1_TMM_long, tissue, cpm, termF_placenta, termM_placenta)

# convert rank_1_TMM into long form
## convert rows into a column
rank_1_TMM_long <- tibble::rownames_to_column(rank_1_TMM, var = "geneid")
## rename, remove "_cpm"
colnames(rank_1_TMM_long) <- gsub("_rank", "", colnames(rank_1_TMM_long))
# convert into long form
rank_1_TMM_long <- tidyr::gather(rank_1_TMM_long, tissue, rank, termF_placenta, termM_placenta)
rank_1_TMM_long_2 <- dplyr::filter(rank_1_TMM_long, rank == "1")

# join the long form tables - nCPM_1_TMM_long and rank_5_TMM
nCPM_rank_1_TMM_long <- dplyr::left_join(rank_1_TMM_long_2, nCPM_1_TMM_long, by = c("geneid", "tissue"))
## rename cpm to nCPM_5
colnames(nCPM_rank_1_TMM_long) <- gsub("cpm", "nCPM_1", colnames(nCPM_rank_1_TMM_long))
```

# create a long df of just placenta_rank = 2 with geneids and ncpms for nCPM(rank=1)/nCPM(rank=2) calculation
```{r}
# create a long df of just placenta_rank == 2  with ensembl IDs and cpms
## create a long df with just the instances where placenta = 2
rank_placenta_long <- filter(rank_1_TMM_long, rank == "2")
## join nCPM_5_TMM_long to rank_placenta_long to create placenta rank 6 cpms table
placenta_2_nCPM <- dplyr::left_join(rank_placenta_long, nCPM_1_TMM_long, by = c("geneid", "tissue"))
### rename so they are unique
colnames(placenta_2_nCPM) <- gsub("cpm", "nCPM_2", colnames(placenta_2_nCPM))
### remove rank and tissue columns
placenta_2_nCPM <- dplyr::select(placenta_2_nCPM, -tissue, -rank)
```

# filter for where nCPM(rank=1)/nCPM(rank=2) > 3
```{r}
# join placenta_2_nCPM and nCPM_rank_5_TMM_long creating a df with both the nCPMs of rank 5 and 6
nCPM1_nCPM2_TMM_male <- dplyr::left_join(nCPM_rank_1_TMM_long, placenta_2_nCPM, by = "geneid")

# create a new column nCPM5/nCPM6
nCPM1_nCPM2_TMM_male$nCPM_ratio <- nCPM1_nCPM2_TMM_male$nCPM_1 / nCPM1_nCPM2_TMM_male$nCPM_2

# select for nCPM_ratio > 3
placenta_m_depleted_TMM <- nCPM1_nCPM2_TMM_male[nCPM1_nCPM2_TMM_male$nCPM_ratio >3,]
```

# tidy and export placenta_f_depleted_TMM
```{r}
# replace column name 'geneid' with 'ensembl'
colnames(placenta_m_depleted_TMM) <- gsub("geneid", "ensembl", colnames(placenta_m_depleted_TMM))

# export placenta_f_depleted_TMM into project rawData for further analysis 
write.csv(placenta_m_depleted_TMM, file = file.path(projectDir, "TMM_placentaM_depleted.csv"))
```

