---
title: "female_enrichment"
output: html_document
date: "2024-08-26"
---

# load libraries and set project directory 
```{r}
# load relevant libraries 
library(ggkegg)
library(tidygraph)
library(scatterpie)
library(dplyr)
library(readxl)
library(readr)
library(edgeR)
library(limma)
library(Glimma)
library(clusterProfiler)
library(enrichplot)
library(stringr)
library(magrittr)
library(RColorBrewer)
library(gprofiler2)
library(forcats)

# set project directory
projectDir <- "/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/readCounts_analysis/rawData"

# set path for depleted gene tables 
import_placentaF_depleted <- file.path(projectDir, "placentaF_depleted_annotated.csv")
import_placentaM_depleted <- file.path(projectDir, "placentaM_depleted_annotated.csv")
```

# import depleted gene tables 
```{r}
# import placenta female depleted table and make it into a data frame
placentaF_depleted <- read_delim(import_placentaF_depleted,
                                 col_names = TRUE,
                                 delim = ",") %>%
  as.data.frame() %>%
  ## tidy placentaF_depleted - remove the "...1" column
  dplyr::select(., -...1)

# import male placenta depleted table and make into a data frame
placentaM_depleted <- read_delim(import_placentaM_depleted,
                                 col_names = TRUE,
                                 delim = ",") %>%
  as.data.frame() %>%
  ## tidy placentaM_depleted - remove the "...1" column
  dplyr::select(., -...1)
```

# GO functional enrichment analysis using gprofiler2 on female placenta depleted genes 
```{r}
# order genes by nCPM_ratio_TMM
placentaF_ordered <- placentaF_depleted[order(placentaF_depleted$nCPM_ratio_TMM, decreasing = TRUE),]

# perform GO enrichment analysis on ordered genes
placentaF_ordered_enriched <- gost(placentaF_ordered$ensembl, organism = "hsapiens",
                                   ordered_query = TRUE,
                                   correction_method = "fdr",
                                   evcodes = TRUE)

# print the enriched terms to visualise initial GO terms before semantic reduction 
head(placentaF_ordered_enriched$result, 30)

# investigate which genes were enriched in fatty acid oxidation related GO terms
go_term_of_interest <- "regulation of fatty acid beta-oxidation"
enriched_row <- placentaF_ordered_enriched$result[placentaF_ordered_enriched$result$term_name == go_term_of_interest, ]
print(enriched_row$intersection)
```

# GO functional enrichment analysis using gprofiler2 on male placenta depleted genes 
```{r}
# order genes by nCPM_ratio_TMM 
placentaM_ordered <- placentaM_depleted[order(placentaM_depleted$nCPM_5_TMM, decreasing = TRUE),]

# perform GO enrichment on ordered genes
placentaM_ordered_enriched <- gost(placentaM_ordered$ensembl, organism = "hsapiens",
                                   ordered_query = TRUE,
                                   correction_method = "fdr",
                                   evcodes = TRUE)

# print the enriched terms to visualise initial GO terms before applying semantic reduction
head(placentaM_ordered_enriched$result, 16)

# investigate which genes were enriched in fatty acid oxidation related GO terms
go_term_of_interest <- "monocarboxylic acid binding"
enriched_row <- placentaM_ordered_enriched$result[placentaM_ordered_enriched$result$term_name == go_term_of_interest, ]
print(enriched_row$intersection)
```

# rrvgo reduce redundancy of (female) human GO terms
```{r}
#BiocManager::install("rrvgo")
library(rrvgo)
library(ComplexHeatmap)

# BP 
## create similarity matrix for GO terms of BP source
simMatrix_BP <- calculateSimMatrix(placentaF_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "BP",
                                method = "Rel")
## set scores
scores <- setNames(-log10(placentaF_ordered_enriched$result$p_value), placentaF_ordered_enriched$result$term_id)
## reduce the number of terms based on the similarity matrix for BP terms
reducedTerms_BP <- reduceSimMatrix(simMatrix_BP,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_BP$source <- "BP"

# MF 
## create similarity matrix for GO terms of MF source
simMatrix_MF <- calculateSimMatrix(placentaF_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "MF",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for MF terms
reducedTerms_MF <- reduceSimMatrix(simMatrix_MF,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_MF$source <- "MF"

# CC 
## create similarity matrix for GO terms of CC source
simMatrix_CC <- calculateSimMatrix(placentaF_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "CC",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for CC terms
reducedTerms_CC <- reduceSimMatrix(simMatrix_CC,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_CC$source <- "CC"


# visualisation of the reduced terms
## heatmap plot
heatmapPlot(simMatrix_BP,
            reducedTerms,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6,
            #show_colnames = FALSE,
            #show_rownames = FALSE,
            fontsize_col = 2,
            fontsize_row = 2)

## scatter plot
scatterPlot(simMatrix, reducedTerms)
## treemap plot
treemapPlot(reducedTerms)

# prelim barplot of parent GO terms and number of associations 
reduced <- dplyr::select(reducedTerms, parent, parentTerm, size)
plot <- ggplot(reduced, aes(x = reorder(parentTerm, -size), y = size, fill = parentTerm)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "GO Term", y = "Number of Associations", title = "Number of Associations per GO Term") +
  theme(legend.text=element_text(size=6)) +
  theme_bw()
plot
```

# rrvgo reduce redundancy of (male) human GO terms
```{r}
# BP 
## create similarity matrix for GO terms of BP source
simMatrixM_BP <- calculateSimMatrix(placentaM_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "BP",
                                method = "Rel")
## set scores
scoresM <- setNames(-log10(placentaM_ordered_enriched$result$p_value), placentaM_ordered_enriched$result$term_id)
## reduce the number of terms based on the similarity matrix for BP terms
reducedTermsM_BP <- reduceSimMatrix(simMatrixM_BP,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_BP$source <- "BP"

# MF 
## create similarity matrix for GO terms of MF source
simMatrixM_MF <- calculateSimMatrix(placentaM_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "MF",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for MF terms
reducedTermsM_MF <- reduceSimMatrix(simMatrixM_MF,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_MF$source <- "MF"

# CC 
## create similarity matrix for GO terms of CC source
simMatrixM_CC <- calculateSimMatrix(placentaM_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "CC",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for CC terms
reducedTermsM_CC <- reduceSimMatrix(simMatrixM_CC,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_CC$source <- "CC"
```

# bar graph visualisation
```{r}
# create a table of reduced terms 
reduced_BP <- dplyr::select(reducedTerms_BP, parent, parentTerm, size, source)
reduced_MF <- dplyr::select(reducedTerms_MF, parent, parentTerm, size, source)
reduced_CC <- dplyr::select(reducedTerms_CC, parent, parentTerm, size, source)
reducedM_BP <- dplyr::select(reducedTermsM_BP, parent, parentTerm, size, source)
reducedM_MF <- dplyr::select(reducedTermsM_MF, parent, parentTerm, size, source)
reducedM_CC <- dplyr::select(reducedTermsM_CC, parent, parentTerm, size, source)

# merge tables
reducedTerms_female <- rbind(reduced_BP, reduced_MF) %>% 
  rbind(., reduced_CC) %>%
  group_by(parentTerm, source) %>%
  summarise(size = sum(size), .groups = 'drop')
## save female reduced term table
write.csv(reducedTerms_female, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/reducedTerms_female.csv"))

reducedTerms_male <- rbind(reducedM_BP, reducedM_MF) %>% 
  rbind(., reducedM_CC) %>%
  group_by(parentTerm, source) %>%
  summarise(size = sum(size), .groups = 'drop')
## save male reduced term table
write.csv(reducedTerms_male, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/reducedTerms_male.csv"))

# create a box plot
reducedTerms_female %>%
  mutate(parentTerm = fct_reorder(parentTerm, desc(size))) %>%
  ggplot( aes(x=parentTerm, y=size, fill = source)) +
  geom_bar(stat="identity", alpha=.7, width=.7) +
  coord_flip() +
  labs(x = "GO Term", y = "Number of Associations", fill = "Source") +
  theme_bw() +
  theme(text = element_text(size = 7),
        axis.text = element_text(color = "black")) +
  scale_y_continuous(expand = c(0, 300)) +
  scale_fill_manual(values = c("#9ECAE1", "#E41A1C", "#B8E186")) 
ggsave("female_go_bar.png", width = 7, height = 7, units = "in")

# male plot 
reducedTerms_male %>%
  mutate(parentTerm = fct_reorder(parentTerm, desc(size))) %>%
  ggplot( aes(x=parentTerm, y=size, fill = source)) +
  geom_bar(stat="identity", alpha=.8, width=.7) +
  coord_flip() +
  labs(x = "GO Term", y = "Number of Associations", fill = "Source") +
  theme_bw() +
  theme(text = element_text(size = 10),
        axis.text = element_text(color = "black")) +
  scale_y_continuous(expand = c(0, 300)) +
  scale_fill_manual(values = c("#9ECAE1", "#E41A1C", "#B8E186")) 
ggsave("male_go_bar.png", width = 7, height = 7, units = "in")
```

# summary visualisation of GO terms
```{r}
# select for summary GO terms
summaryTerms_female <- dplyr::filter(reducedTerms_female, parentTerm %in% c("regulation of anatomical structure morphogenesis", "positive regulation of phosphorus metabolic process", "synapse organization", "smooth muscle tissue development", "regulation of fatty acid oxidation", "fatty acid oxidation", "response to folic acid", "microtubule severing ATPase activity", "integrin binding", "trans-1,2-dihydrobenzene-1,2-diol dehydrogenase activity"))

## add a column defning sex
summaryTerms_female$tissue <- "Mid-gestation Placenta (Female)"

# select for summary GO terms
summaryTerms_male <- dplyr::filter(reducedTerms_male, parentTerm %in% c("vitamin B6 binding", "sulfur compound binding", "integrin binding", "microtubule severing ATPase activity", "trans-1,2-dihydrobenzene-1,2-diol dehydrogenase activity", "thrombin-activated receptor activity")) 

## add a column defining the sex
summaryTerms_male$tissue <- "Mid-gestation Placenta (Male)"

# combine the symmaryTerms dfs
summaryTerms <- rbind(summaryTerms_female, summaryTerms_male)

# create a dot plot
summaryTerms %>%
  ggplot( aes(x=tissue, y=parentTerm, color = source, size = size)) +
  geom_point(stat="identity", show.legend = TRUE) +
  labs(x = "Tissue", y = "Parent GO Term", color = "Source", size = "Number of Associations") +
  theme_bw() +
  theme(text = element_text(size = 10, family = "Times New Roman"),
        axis.text = element_text(color = "black"),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        legend.text = element_text(size = 10)) +
  scale_color_manual(values = c("#9ECAE1", "#B8E186"))
ggsave("fetal_go_point.png", width = 8.5, height = 7, units = "in")
```


# GO analysis of female and male uniquely depleted gene sets
```{r}
# load in data 
## female unique depletion
female_depletion <- read_delim(file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/tissue_depletion/rawData/unique_female_depletion.csv"),
                               col_names = TRUE,
                               delim = ",") %>% 
  dplyr::select(., -...1)

## male unique depletion
male_depletion <- read_delim(file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/tissue_depletion/rawData/unique_male_depletion.csv"),
                               col_names = TRUE,
                               delim = ",") %>% 
  dplyr::select(., -...1)


# gprofiler2 GO analysis
## female
## perform GO enrichment analysis on genes
female_enriched <- gost(female_depletion$ensembl, organism = "hsapiens",
                                   ordered_query = TRUE,
                                   correction_method = "fdr")
## print the enriched terms to visualise initial GO terms before semantic reduction 
print(female_enriched$result)


## male 
## perform GO enrichment analysis on genes
male_enriched <- gost(male_depletion$ensembl, organism = "hsapiens",
                                   ordered_query = TRUE,
                                   correction_method = "fdr")
## print the enriched terms to visualise initial GO terms before semantic reduction 
head(male_enriched$result, 30)

```

# rrvgo reduce redundancy of unique female human GO terms
```{r}
#BiocManager::install("rrvgo")
library(rrvgo)
library(ComplexHeatmap)

# BP 
## create similarity matrix for GO terms of BP source
simMatrix_BP <- calculateSimMatrix(female_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "BP",
                                method = "Rel")
## set scores
scores <- setNames(-log10(female_enriched$result$p_value), female_enriched$result$term_id)
## reduce the number of terms based on the similarity matrix for BP terms
reducedTerms_BP <- reduceSimMatrix(simMatrix_BP,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_BP$source <- "BP"

# MF 
## create similarity matrix for GO terms of MF source
simMatrix_MF <- calculateSimMatrix(female_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "MF",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for MF terms
reducedTerms_MF <- reduceSimMatrix(simMatrix_MF,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_MF$source <- "MF"

# CC 
## create similarity matrix for GO terms of CC source
simMatrix_CC <- calculateSimMatrix(female_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "CC",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for CC terms
reducedTerms_CC <- reduceSimMatrix(simMatrix_CC,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_CC$source <- "CC"

# create a table of reduced terms 
reduced_BP <- dplyr::select(reducedTerms_BP, parent, parentTerm, size, source)
reduced_MF <- dplyr::select(reducedTerms_MF, parent, parentTerm, size, source)
reduced_CC <- dplyr::select(reducedTerms_CC, parent, parentTerm, size, source)
reducedM_BP <- dplyr::select(reducedTermsM_BP, parent, parentTerm, size, source)
reducedM_MF <- dplyr::select(reducedTermsM_MF, parent, parentTerm, size, source)
reducedM_CC <- dplyr::select(reducedTermsM_CC, parent, parentTerm, size, source)

# merge tables
reduced_combined <- rbind(reduced_BP, reduced_MF) %>% 
  rbind(., reduced_CC) %>%
  group_by(parentTerm, source) %>%
  summarise(size = sum(size), .groups = 'drop')

male <- rbind(reducedM_BP, reducedM_MF) %>% 
  rbind(., reducedM_CC) %>%
  group_by(parentTerm, source) %>%
  summarise(size = sum(size), .groups = 'drop')

# create a box plot
  ggplot(reduced_combined, aes(x=parentTerm, y=size, fill = source)) +
  geom_bar(stat="identity", alpha=.8, width=.7) +
  coord_flip() +
  labs(x = "GO Term", y = "Number of Associations") +
  theme_bw() +
  theme(text = element_text(size = 10)) +
  scale_y_continuous(expand = c(0, 300))
```

# rrvgo reduce redundancy of unique male human GO terms
```{r}
# BP 
## create similarity matrix for GO terms of BP source
simMatrixM_BP <- calculateSimMatrix(male_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "BP",
                                method = "Rel")
## set scores
scoresM <- setNames(-log10(male_enriched$result$p_value), male_enriched$result$term_id)
## reduce the number of terms based on the similarity matrix for BP terms
reducedTermsM_BP <- reduceSimMatrix(simMatrixM_BP,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_BP$source <- "BP"

# MF 
## create similarity matrix for GO terms of MF source
simMatrixM_MF <- calculateSimMatrix(male_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "MF",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for MF terms
reducedTermsM_MF <- reduceSimMatrix(simMatrixM_MF,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_MF$source <- "MF"

# CC 
## create similarity matrix for GO terms of CC source
simMatrixM_CC <- calculateSimMatrix(male_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "CC",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for CC terms
reducedTermsM_CC <- reduceSimMatrix(simMatrixM_CC,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_CC$source <- "CC"

# create a table of reduced terms 
reducedM_BP <- dplyr::select(reducedTermsM_BP, parent, parentTerm, size, source)
reducedM_MF <- dplyr::select(reducedTermsM_MF, parent, parentTerm, size, source)
reducedM_CC <- dplyr::select(reducedTermsM_CC, parent, parentTerm, size, source)

# merge tables
male <- rbind(reducedM_BP, reducedM_MF) %>% 
  rbind(., reducedM_CC) %>%
  group_by(parentTerm, source) %>%
  summarise(size = sum(size), .groups = 'drop')
  
# male plot 
ggplot(male, aes(x=parentTerm, y=size, fill = source)) +
  geom_bar(stat="identity", alpha=.8, width=.7) +
  coord_flip() +
  labs(x = "GO Term", y = "Number of Associations") +
  theme_bw() +
  theme(text = element_text(size = 10)) +
  scale_y_continuous(expand = c(0, 300))
```

# export reduced go lists
```{r}
# female depletion unique list
write.csv(reduced_combined, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/unique depletions/female_unique_go_terms.csv"))

# male depletion unique list
write.csv(male, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/unique depletions/male_unique_go_terms.csv"))

## female comparison original go terms
original_female <- female_enriched$result %>% 
  as.data.frame() %>% 
  dplyr::select(., term_name, source)
write.csv(original_female, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/unique depletions/female_unique_original_go.csv"))

## male comparison original go terms
original_male <- male_enriched$result %>% 
  as.data.frame() %>% 
  dplyr::select(., term_name, source)
write.csv(original_male, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/unique depletions/male_unique_original_go.csv"))

## combined depletion list
write.csv(reduced_all, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/unique depletions/fetal_combined_go_terms.csv"))

## combined comparison original go terms
original_all <- combined_enriched$result %>% 
  as.data.frame() %>% 
  dplyr::select(., term_name, source)
write.csv(original_all, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/unique depletions/fetal_combined_original_go.csv"))
```

# overlapping gene depletion GO analysis
```{r}
# created combined df
combined_depletion <- dplyr::inner_join(placentaF_depleted, placentaM_depleted, by = join_by("ensembl"))
## export overlapping genes for scatter plot
write.csv(combined_depletion, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/readCounts_analysis/rawData/overlapping_depletion.csv"))

# GO analysis on combined depletion
## perform GO enrichment analysis on genes
combined_enriched <- gost(combined_depletion$ensembl, organism = "hsapiens",
                                   ordered_query = TRUE,
                                   correction_method = "fdr")
## print the enriched terms to visualise initial GO terms before semantic reduction 
print(combined_enriched$result)
```

# rrvgo reduce redundancy of combined depleted human GO terms
```{r}
# BP 
## create similarity matrix for GO terms of BP source
simMatrixC_BP <- calculateSimMatrix(combined_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "BP",
                                method = "Rel")
## set scores
scoresC <- setNames(-log10(combined_enriched$result$p_value), combined_enriched$result$term_id)
## reduce the number of terms based on the similarity matrix for BP terms
reducedTermsC_BP <- reduceSimMatrix(simMatrixC_BP,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresC)
## add the source to table
reducedTermsC_BP$source <- "BP"

# MF 
## create similarity matrix for GO terms of MF source
simMatrixC_MF <- calculateSimMatrix(combined_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "MF",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for MF terms
reducedTermsC_MF <- reduceSimMatrix(simMatrixC_MF,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresC)
## add the source to table
reducedTermsC_MF$source <- "MF"

# CC 
## create similarity matrix for GO terms of CC source
simMatrixC_CC <- calculateSimMatrix(combined_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "CC",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for CC terms
reducedTermsC_CC <- reduceSimMatrix(simMatrixC_CC,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresC)
## add the source to table
reducedTermsC_CC$source <- "CC"

# create a table of reduced terms 
reducedC_BP <- dplyr::select(reducedTermsC_BP, parent, parentTerm, size, source)
reducedC_MF <- dplyr::select(reducedTermsC_MF, parent, parentTerm, size, source)
reducedC_CC <- dplyr::select(reducedTermsC_CC, parent, parentTerm, size, source)

# merge tables
reduced_all <- rbind(reducedC_BP, reducedC_MF) %>% 
  rbind(., reducedC_CC) %>%
  group_by(parentTerm, source) %>%
  summarise(size = sum(size), .groups = 'drop')

# create a box plot
  ggplot(reduced_all, aes(x=parentTerm, y=size, fill = source)) +
  geom_bar(stat="identity", alpha=.8, width=.7) +
  coord_flip() +
  labs(x = "GO Term", y = "Number of Associations") +
  theme_bw() +
  theme(text = element_text(size = 10)) +
  scale_y_continuous(expand = c(0, 300))
```

