---
title: "mid_rrvgo"
output: html_document
date: "2024-08-27"
---

# load libraries and set project directory 
```{r}
# load relevant libraries 
library(ggkegg)
library(tidygraph)
library(scatterpie)
library(dplyr)
library(readxl)
library(readr)
library(edgeR)
library(limma)
library(Glimma)
library(clusterProfiler)
library(enrichplot)
library(stringr)
library(magrittr)
library(RColorBrewer)
library(gprofiler2)
library(forcats)

# set project directory
projectDir <- "/Users/taylaalbertini/Desktop/R studio/honours_project/oxygen_threshold/rawData"

# set path for depleted gene tables 
import_placentaF_depleted <- file.path(projectDir, "mid_placentaF_depleted_annotated.csv")
import_placentaM_depleted <- file.path(projectDir, "mid_placentaM_depleted_annotated.csv")
```

# import depleted gene tables 
```{r}
# import placenta female depleted table and make it into a data frame
placentaF_depleted <- read_delim(import_placentaF_depleted,
                                 col_names = TRUE,
                                 delim = ",") %>%
  as.data.frame() %>%
  ## tidy placentaF_depleted - remove the "...1" column
  dplyr::select(., -...1)

# import male placenta depleted table and make into a data frame
placentaM_depleted <- read_delim(import_placentaM_depleted,
                                 col_names = TRUE,
                                 delim = ",") %>%
  as.data.frame() %>%
  ## tidy placentaM_depleted - remove the "...1" column
  dplyr::select(., -...1)
```

# GO functional enrichment analysis using gprofiler2 on female placenta depleted genes 
```{r}
# order genes by nCPM_ratio_TMM
placentaF_ordered <- placentaF_depleted[order(placentaF_depleted$nCPM_ratio_TMM, decreasing = TRUE),]

# perform GO enrichment analysis on ordered genes
placentaF_ordered_enriched <- gost(placentaF_ordered$ensembl, organism = "hsapiens",
                                   ordered_query = TRUE,
                                   correction_method = "fdr",
                                   evcodes = TRUE)

# print the enriched terms to visualise initial GO terms before semantic reduction 
head(placentaF_ordered_enriched$result, 30)

# investigate which genes were enriched in GO terms
go_term_of_interest <- "lipoprotein lipid oxidation"
enriched_row <- placentaF_ordered_enriched$result[placentaF_ordered_enriched$result$term_name == go_term_of_interest, ]
print(enriched_row$intersection)

# investigating a gene of interest for associated GO terms
# Define the gene of interest
gene_of_interest <- "ENSG00000143627"
# Filter the enriched results to find rows where the gene of interest is in the intersection
associated_go_terms <- placentaF_ordered_enriched$result[
  grepl(gene_of_interest, placentaF_ordered_enriched$result$intersection), ]
# Print the GO terms associated with the gene of interest
print(associated_go_terms$term_name)

# Print the found genes
print(found_genes)
```

# GO functional enrichment analysis using gprofiler2 on male placenta depleted genes 
```{r}
# order genes by nCPM_ratio_TMM 
placentaM_ordered <- placentaM_depleted[order(placentaM_depleted$nCPM_ratio_TMM, decreasing = TRUE),]

# perform GO enrichment on ordered genes
placentaM_ordered_enriched <- gost(placentaM_ordered$ensembl, organism = "hsapiens",
                                   ordered_query = TRUE,
                                   correction_method = "fdr",
                                   evcodes = TRUE)

# print the enriched terms to visualise initial GO terms before applying semantic reduction
head(placentaM_ordered_enriched$result, 16)

# investigate which genes were enriched in GO terms
go_term_of_interest <- "carbohydrate metabolic process"
enriched_row <- placentaM_ordered_enriched$result[placentaM_ordered_enriched$result$term_name == go_term_of_interest, ]
print(enriched_row$intersection)

# determine if carbohydrate metabolic process related genes are present in placentalM_depleted
# Define the list of gene IDs (as a character vector)
gene_ids_of_interestM <- c("ENSG00000143627","ENSG00000140297","ENSG00000167080","ENSG00000143365","ENSG00000156510","ENSG00000149328","ENSG00000138079","ENSG00000160180")

# Assuming your dataframe is called 'df' and the column with gene IDs is named 'gene_column'
# Modify 'df' and 'gene_column' to match your actual dataframe and column names
found_genesM <- placentaM_depleted[placentaM_depleted$ensembl %in% gene_ids_of_interestM, ]

# Print the found genes
print(found_genesM)
```

# rrvgo reduce redundancy of (mid female) human GO terms
```{r}
#BiocManager::install("rrvgo")
library(rrvgo)
library(ComplexHeatmap)

# BP 
## create similarity matrix for GO terms of BP source
simMatrix_BP <- calculateSimMatrix(placentaF_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "BP",
                                method = "Rel")
## set scores
scores <- setNames(-log10(placentaF_ordered_enriched$result$p_value), placentaF_ordered_enriched$result$term_id)
## reduce the number of terms based on the similarity matrix for BP terms
reducedTerms_BP <- reduceSimMatrix(simMatrix_BP,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_BP$source <- "BP"

# MF 
## create similarity matrix for GO terms of MF source
simMatrix_MF <- calculateSimMatrix(placentaF_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "MF",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for MF terms
reducedTerms_MF <- reduceSimMatrix(simMatrix_MF,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_MF$source <- "MF"

# CC 
## create similarity matrix for GO terms of CC source
simMatrix_CC <- calculateSimMatrix(placentaF_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "CC",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for CC terms
reducedTerms_CC <- reduceSimMatrix(simMatrix_CC,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_CC$source <- "CC"
```

# rrvgo reduce redundancy of (mid male) human GO terms
```{r}
# BP 
## create similarity matrix for GO terms of BP source
simMatrixM_BP <- calculateSimMatrix(placentaM_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "BP",
                                method = "Rel")
## set scores
scoresM <- setNames(-log10(placentaM_ordered_enriched$result$p_value), placentaM_ordered_enriched$result$term_id)
## reduce the number of terms based on the similarity matrix for BP terms
reducedTermsM_BP <- reduceSimMatrix(simMatrixM_BP,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_BP$source <- "BP"

# MF 
## create similarity matrix for GO terms of MF source
simMatrixM_MF <- calculateSimMatrix(placentaM_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "MF",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for MF terms
reducedTermsM_MF <- reduceSimMatrix(simMatrixM_MF,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_MF$source <- "MF"

# CC 
## create similarity matrix for GO terms of CC source
simMatrixM_CC <- calculateSimMatrix(placentaM_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "CC",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for CC terms
reducedTermsM_CC <- reduceSimMatrix(simMatrixM_CC,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_CC$source <- "CC"
```

# bar graph visualisation
```{r}
# create a table of reduced terms 
reduced_BP <- dplyr::select(reducedTerms_BP, parent, parentTerm, size, source)
reduced_MF <- dplyr::select(reducedTerms_MF, parent, parentTerm, size, source)
reduced_CC <- dplyr::select(reducedTerms_CC, parent, parentTerm, size, source)
reducedM_BP <- dplyr::select(reducedTermsM_BP, parent, parentTerm, size, source)
reducedM_MF <- dplyr::select(reducedTermsM_MF, parent, parentTerm, size, source)
reducedM_CC <- dplyr::select(reducedTermsM_CC, parent, parentTerm, size, source)

# merge tables
reducedTerms_midF <- rbind(reduced_BP, reduced_MF) %>% 
  rbind(., reduced_CC) %>%
  group_by(parentTerm, source) %>%
  summarise(size = sum(size), .groups = 'drop')
## export reduced mid female terms
write.csv(reducedTerms_midF, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/raw_data/reducedTerms_midF.csv"))

reducedTerms_midM <- rbind(reducedM_BP, reducedM_MF) %>% 
  rbind(., reducedM_CC) %>%
  group_by(parentTerm, source) %>%
  summarise(size = sum(size), .groups = 'drop')
## export reduced mid male terms
write.csv(reducedTerms_midM, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/raw_data/reducedTerms_midM.csv"))

# create a box plot
reducedTerms_midF %>%
  mutate(parentTerm = fct_reorder(parentTerm, desc(size))) %>%
  ggplot( aes(x=parentTerm, y=size, fill = source)) +
  geom_bar(stat="identity", alpha=.7, width=.7) +
  coord_flip() +
  labs(x = "GO Term", y = "Number of Associations", fill = "Source") +
  theme_bw() +
  theme(text = element_text(size = 6)) +
  scale_y_continuous(expand = c(0, 300)) +
  scale_fill_manual(values = c("#9ECAE1", "#E41A1C", "#B8E186")) 
ggsave("midF_go_bar.png", width = 7, height = 7, units = "in")

# male plot 
reducedTerms_midM %>%
  mutate(parentTerm = fct_reorder(parentTerm, desc(size))) %>%
  ggplot( aes(x=parentTerm, y=size, fill = source)) +
  geom_bar(stat="identity", alpha=.7, width=.7) +
  coord_flip() +
  labs(x = "GO Term", y = "Number of Associations", fill = "Source") +
  theme_bw() +
  theme(text = element_text(size = 6)) +
  scale_y_continuous(expand = c(0, 300)) +
  scale_fill_manual(values = c("#9ECAE1", "#E41A1C", "#B8E186")) 
ggsave("midM_go_bar.png", width = 7, height = 7, units = "in")
```

# summary visualisation of GO terms
```{r}
# select for summary GO terms
summaryTerms_female <- dplyr::filter(reducedTerms_midF, parentTerm %in% c("multicellular organismal process", "carbohydrate metabolic process", "G-protein alpha-subunit binding", "insulin-like growth factor II binding", "tissue development", "response to stimulus", "system development", "estrogen response element binding", "positive regulation of cytokine production"))

## add a column defning sex
summaryTerms_female$tissue <- "Mid-gestation Placenta (Female)"

# select for summary GO terms
summaryTerms_male <- dplyr::filter(reducedTerms_midM, parentTerm %in% c("cell differentiation", "anatomical structure morphogenesis", "multicellular organismal process", "follicle-stimulating hormone secretion", "actin filament capping", "lipoprotein lipid oxidation", "reactive oxygen species metabolic process", "carbohydrate metabolic process", "G-protein alpha-subunit binding", "BMP binding", "insulin-like growth factor binding", "tissue development", "response to stimulus")) 

## add a column defining the sex
summaryTerms_male$tissue <- "Mid-gestation Placenta (Male)"

# combine the symmaryTerms dfs
summaryTerms <- rbind(summaryTerms_female, summaryTerms_male)

# create a dot plot
summaryTerms %>%
  ggplot( aes(x=tissue, y=parentTerm, color = source, size = size)) +
  geom_point(stat="identity", show.legend = TRUE) +
  labs(x = "Tissue", y = "Parent GO Term", color = "Source", size = "Number of Associations") +
  theme_bw() +
  theme(text = element_text(size = 10, family = "Times New Roman"),
        axis.text = element_text(color = "black"),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        legend.text = element_text(size = 10)) +
  scale_color_manual(values = c("#9ECAE1", "#B8E186"))
ggsave("mid_go_point.png", width = 8.5, height = 7, units = "in")
```
