---
title: "term_rrvgo"
output: html_document
date: "2024-08-27"
---

# load libraries
```{r}
library(stringr)
library(magrittr)
library(ggplot2)
library(RColorBrewer)
library(ggkegg)
library(tidygraph)
library(scatterpie)
library(dplyr)
library(readxl)
library(readr)
library(edgeR)
library(limma)
library(Glimma)
library(clusterProfiler)
library(enrichplot)
library(gprofiler2)

# set project directory
projectDir <- "/Users/taylaalbertini/Desktop/R studio/honours_project/termPlacenta_comparison/rawData"
import_placentaF_depleted <- file.path(projectDir, "term_placentaF_depleted_annotated.csv")
import_placnetaM_depleted <- file.path(projectDir, "term_placentaM_depleted_annotated.csv")
```

# import files
```{r}
# import placenta female depleted table and make it into a data frame
placentaF_depleted <- read_delim(import_placentaF_depleted,
                                 col_names = TRUE,
                                 delim = ",") %>%
  as.data.frame() 
  ## tidy placentaF_depleted - remove the "...1" column
  placentaF_depleted <- dplyr::select(placentaF_depleted, -...1)

# import male placenta depleted table and make into a data frame
placentaM_depleted <- read_delim(import_placnetaM_depleted,
                                 col_names = TRUE,
                                 delim = ",") %>%
  as.data.frame()
  ## tidy placentaM_depleted - remove the "...1" column
  placentaM_depleted <- dplyr::select(placentaM_depleted, -...1)
```

# GO functional enrichment analysis using gprofiler2 on female placenta depleted genes 
```{r}
# order genes by nCPM_ratio_TMM
placentaF_ordered <- placentaF_depleted[order(placentaF_depleted$nCPM_ratio_TMM, decreasing = TRUE),]

# perform GO enrichment analysis on ordered genes
placentaF_ordered_enriched <- gost(placentaF_ordered$ensembl, organism = "hsapiens",
                                   ordered_query = TRUE,
                                   correction_method = "fdr",
                                   evcodes = TRUE)

# print the enriched terms to visualise initial GO terms before semantic reduction 
head(placentaF_ordered_enriched$result, 30)

# investigate which genes were enriched in GO terms
go_term_of_interest <- "phospholipid metabolic process"
enriched_row <- placentaF_ordered_enriched$result[placentaF_ordered_enriched$result$term_name == go_term_of_interest, ]
print(enriched_row$intersection)
```

# GO functional enrichment analysis using gprofiler2 on male placenta depleted genes 
```{r}
# order genes by nCPM_ratio_TMM 
placentaM_ordered <- placentaM_depleted[order(placentaM_depleted$nCPM_ratio_TMM, decreasing = TRUE),]

# perform GO enrichment on ordered genes
placentaM_ordered_enriched <- gost(placentaM_ordered$ensembl, organism = "hsapiens",
                                   ordered_query = TRUE,
                                   correction_method = "fdr",
                                   evcodes = TRUE)

# print the enriched terms to visualise initial GO terms before applying semantic reduction
head(placentaM_ordered_enriched$result, 16)

# investigate which genes were enriched in GO terms
go_term_of_interest <- "phospholipid metabolic process"
enriched_row <- placentaF_ordered_enriched$result[placentaF_ordered_enriched$result$term_name == go_term_of_interest, ]
print(enriched_row$intersection)
```

# rrvgo reduce redundancy of (term female) human GO terms
```{r}
#BiocManager::install("rrvgo")
library(rrvgo)
library(ComplexHeatmap)

# BP 
## create similarity matrix for GO terms of BP source
simMatrix_BP <- calculateSimMatrix(placentaF_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "BP",
                                method = "Rel")
## set scores
scores <- setNames(-log10(placentaF_ordered_enriched$result$p_value), placentaF_ordered_enriched$result$term_id)
## reduce the number of terms based on the similarity matrix for BP terms
reducedTerms_BP <- reduceSimMatrix(simMatrix_BP,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_BP$source <- "BP"

# MF 
## create similarity matrix for GO terms of MF source
simMatrix_MF <- calculateSimMatrix(placentaF_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "MF",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for MF terms
reducedTerms_MF <- reduceSimMatrix(simMatrix_MF,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_MF$source <- "MF"

# CC 
## create similarity matrix for GO terms of CC source
simMatrix_CC <- calculateSimMatrix(placentaF_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "CC",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for CC terms
reducedTerms_CC <- reduceSimMatrix(simMatrix_CC,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scores)
## add the source to table
reducedTerms_CC$source <- "CC"
```

# rrvgo reduce redundancy of (term male) human GO terms
```{r}
# BP 
## create similarity matrix for GO terms of BP source
simMatrixM_BP <- calculateSimMatrix(placentaM_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "BP",
                                method = "Rel")
## set scores
scoresM <- setNames(-log10(placentaM_ordered_enriched$result$p_value), placentaM_ordered_enriched$result$term_id)
## reduce the number of terms based on the similarity matrix for BP terms
reducedTermsM_BP <- reduceSimMatrix(simMatrixM_BP,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_BP$source <- "BP"

# MF 
## create similarity matrix for GO terms of MF source
simMatrixM_MF <- calculateSimMatrix(placentaM_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "MF",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for MF terms
reducedTermsM_MF <- reduceSimMatrix(simMatrixM_MF,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_MF$source <- "MF"

# CC 
## create similarity matrix for GO terms of CC source
simMatrixM_CC <- calculateSimMatrix(placentaM_ordered_enriched$result$term_id,
                                orgdb = "org.Hs.eg.db",
                                ont = "CC",
                                method = "Rel")
## reduce the number of terms based on the similarity matrix for CC terms
reducedTermsM_CC <- reduceSimMatrix(simMatrixM_CC,
                                threshold = 0.85,
                                orgdb = "org.Hs.eg.db",
                                scoresM)
## add the source to table
reducedTermsM_CC$source <- "CC"
```

# bar graph visualisation
```{r}
# create a table of reduced terms 
reduced_BP <- dplyr::select(reducedTerms_BP, parent, parentTerm, size, source)
reduced_MF <- dplyr::select(reducedTerms_MF, parent, parentTerm, size, source)
reduced_CC <- dplyr::select(reducedTerms_CC, parent, parentTerm, size, source)
reducedM_BP <- dplyr::select(reducedTermsM_BP, parent, parentTerm, size, source)
reducedM_MF <- dplyr::select(reducedTermsM_MF, parent, parentTerm, size, source)
reducedM_CC <- dplyr::select(reducedTermsM_CC, parent, parentTerm, size, source)

# merge tables
reducedTerms_termF <- rbind(reduced_BP, reduced_MF) %>% 
  rbind(., reduced_CC) %>%
  group_by(parentTerm, source) %>%
  summarise(size = sum(size), .groups = 'drop')
## export reduced mid female terms
write.csv(reducedTerms_termF, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/raw_data/reducedTerms_termF.csv"))

reducedTerms_termM <- rbind(reducedM_BP, reducedM_MF) %>% 
  rbind(., reducedM_CC) %>%
  group_by(parentTerm, source) %>%
  summarise(size = sum(size), .groups = 'drop')
## export reduced mid male terms
write.csv(reducedTerms_termM, file = file.path("/Users/taylaalbertini/Desktop/R studio/honours_project/fetal_comparison/enrichmentAnalysis_fetal.Rproj/raw_data/reducedTerms_termM.csv"))

# create a box plot
reducedTerms_termF %>%
  mutate(parentTerm = fct_reorder(parentTerm, desc(size))) %>%
  ggplot( aes(x=parentTerm, y=size, fill = source)) +
  geom_bar(stat="identity", alpha=.8, width=.7) +
  coord_flip() +
  labs(x = "GO Term", y = "Number of Associations", fill = "Source") +
  theme_bw() +
  theme(text = element_text(size = 5)) +
  scale_y_continuous(expand = c(0, 300)) +
  scale_fill_manual(values = c("#9ECAE1", "#E41A1C", "#B8E186")) 
ggsave("termF_go_bar.png", width = 7, height = 7, units = "in")

# male plot 
reducedTerms_termM %>%
  mutate(parentTerm = fct_reorder(parentTerm, desc(size))) %>%
  ggplot( aes(x=parentTerm, y=size, fill = source)) +
  geom_bar(stat="identity", alpha=.8, width=.7) +
  coord_flip() +
  labs(x = "GO Term", y = "Number of Associations", fill = "Source") +
  theme_bw() +
  theme(text = element_text(size = 5)) +
  scale_y_continuous(expand = c(0, 300)) +
  scale_fill_manual(values = c("#9ECAE1", "#E41A1C", "#B8E186")) 
ggsave("termM_go_bar.png", width = 7, height = 7, units = "in")
```

# summary visualisation of GO terms
```{r}
# select for summary GO terms
summaryTerms_female <- dplyr::filter(reducedTerms_termF, parentTerm %in% c("developmental process", "fat-soluble vitamin catabolic process", "cholesterol transfer activity", "cholesterol metabolic process", "phospholipid metabolic process", "high-density lipoprotein particle clearance", "phosphatidylcholine biosynthetic process", "fructokinase activity", "ipid binding", "negative regulation of gluconeogenesis", "phospholipid metabolic process", "endoplasmic reticulum lumen"))

## add a column defning sex
summaryTerms_female$tissue <- "Mid-gestation Placenta (Female)"

# select for summary GO terms
summaryTerms_male <- dplyr::filter(reducedTerms_termM, parentTerm %in% c("developmental process", "fat-soluble vitamin catabolic process", "cholesterol transfer activity", "cholesterol metabolic process", "phospholipid metabolic process", "high-density lipoprotein particle clearance", "phosphatidylcholine biosynthetic process", "positive regulation of carbohydrate metabolic process", "endoplasmic reticulum lumen")) 

## add a column defining the sex
summaryTerms_male$tissue <- "Mid-gestation Placenta (Male)"

# combine the symmaryTerms dfs
summaryTerms <- rbind(summaryTerms_female, summaryTerms_male)

# create a dot plot
summaryTerms %>%
  ggplot( aes(x=tissue, y=parentTerm, color = source, size = size)) +
  geom_point(stat="identity", show.legend = TRUE) +
  labs(x = "Tissue", y = "Parent GO Term", color = "Source", size = "Number of Associations") +
  theme_bw() +
  theme(text = element_text(size = 10, family = "Times New Roman"),
        axis.text = element_text(color = "black"),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_blank(),
        legend.text = element_text(size = 10)) +
  scale_color_manual(values = c("#9ECAE1", "#E41A1C",  "#B8E186"))
ggsave("term_go_point.png", width = 8.5, height = 7, units = "in")
```
